---
title: "The Impact of ChatGPT news on MSFT"
author: "Ansam Zedan, Daniel Wullschleger"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Insert introduction here

First we get the MSFT price data via yahoo API and we do an Augmented Dickey-Fuller (ADF) Test to see whether the data is already stationary.

```{r echo=FALSE}
if (!require("vars")) install.packages("vars", dependencies=TRUE)
if (!require("tseries")) install.packages("tseries", dependencies=TRUE)
if (!require("quantmod")) install.packages("quantmod", dependencies=TRUE)

library(vars)
library(tseries)
library(quantmod)

MSFT <- getSymbols.yahoo("MSFT", from="2022-01-01", to="2023-11-18", periodicity = "daily", auto.assign=FALSE)[,6]

adf.test(MSFT) 
```

We see that based on the p-value of the ADF Test, that the price data itself is non-stationary.
So we a apply a log-transform with subsequent differencing to calculate continuous returns on MSFT.
These should behave stationary.

```{r pressure, echo=FALSE}
MSFT_returns <- na.omit(diff(log(MSFT)))
adf.test(MSFT_returns)
```
Now, that we deal with stationary data we can begin with our analysis. 

First we include the ChatGPT News Time Series to then create a Vector Autoregression model as we want to find out, whether ChatGPT News are a causal driver for MSFT price returns.

We then do a coefficient test, to get more information on the statistical significance of the coefficients of our VAR Model.




```{r pressure, echo=FALSE}
coeftest(VAR_est)
MSFT_VAR <- VAR(cbind(MSFT_returns, ChatGPT), ic="AIC", lag.max = 24)

```

As not all coeffcients are signifcant, we may do a causality test. This way, we can find out, if ChatGPT News are in fact a causal driver for MSFT prices



```{r pressure, echo=FALSE}

causality(MSFT_VAR, cause="ChatGPT")["Granger"] 
causality(VAR_est, cause="Inflation_CH")["Granger"] 

```

Lastly, it would be interesting to find out how ChatGPT News actually influence the MSFT price returns. So that we may estimate long  such news do have effect on the price.

For this we simulate shocks to the system


```{r pressure, echo=FALSE}

plot(irf(VAR_est, impulse="ChatGPT", response="MSFT_returns"))


```